---
title: "An Analysis of NBA Player Performance Using Principal Component Analysis"
author: "Jordan Iserman, Jeshurun Moses, and Harsha Pola"
format: revealjs
editor: visual
---

## Introduction

- What is principal component analysis (PCA)?
- What dataset is being used?


## Methods: Linear Combinations

```{r,echo=FALSE, include=FALSE}
library(readxl)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(gtable)
library(RColorBrewer)
library(factoextra)
library(imputeTS)
library(psych)
library(ggcorrplot)
library(corrplot)
library(lares)
library(pvclust)


# Read the excel file into a data frame
player.df <- read.csv("/Users/theji/Documents/Advanced Statistical Modelling/nba.csv", header = TRUE)
# Check the structure of the data frame
#str(player.df)
nba = player.df[,4:ncol(player.df)]
nba = as.data.frame(sapply(nba, as.numeric ))
names = player.df[,1]
nba_data <- na_mean(nba)
pca = prcomp(nba_data, scale=T)
features <- nba_data[, c(
  "AGE", "GP", "MPG", "MINpct", "USGpct", "TOpct",
  "FTA", "Ftpct", "TwoPA", "pct2P", "ThreePA", "pct3P",
  "eFGpct", "Tspct", "PPG", "RPG", "TRBpct", "APG", "ASTpct",
  "SPG", "BPG", "TOPG", "VI", "ORTG", "DRTG"
)]

# Standardize the features (mean=0, sd=1)
standardized_features <- scale(features)

# Calculate the correlation matrix
correlation_matrix <- cor(standardized_features)

corrplot(correlation_matrix, method = "color", type = "upper", tl.cex = 0.7)
```

- Principal component analysis reduces the dimensionality of data sets
- Multiple dimensions can be turned into a linear combination
- The linear combination is a unit vector, or a vector of length 1
- Multiple principal components can be represented on a coordinate plane


## Data: Correlation Matrix

```{r}
corrplot(correlation_matrix, method = "color", type = "upper", tl.cex = 0.7)
```

## Data: Best Performers

```{r}
#TOP 10 PLAYERS BY POINTS PER GAME
# Sorting players by points per game
top_points_players <- player.df[order(player.df$PPG, decreasing = TRUE), ]

# Selecting the top 10 players
top_10_points_players <- top_points_players[1:10, ]

# Creating a bar plot for the top 10 players by points per game
ggplot(top_10_points_players, aes(x = reorder(NAME, -PPG), y = PPG, fill = NAME)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Top 10 Players by Points Per Game", x = "Player", y = "Points Per Game") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Data: Best Performers

```{r}
##TOP 10 PLAYERS BY 3-POINTERS PER GAME

# Sorting players by 3-point percentage
top_3p_players <- player.df[order(player.df$pct3P, decreasing = TRUE), ]
# Selecting the top 10 players
top_10_3p_players <- top_3p_players[1:10, ]
# Creating a bar plot for the top 10 players by 3-point percentage
ggplot(top_10_3p_players, aes(x = reorder(NAME, -pct3P), y = pct3P, fill = NAME)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Top 10 Players by 3-Pointers Percentage", x = "Player", y = "3-Point Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Data: Best Performers

```{r}
##TOP 10 PLAYERS BY ASSISTS PER GAME

# Sorting players by assists
top_assists_players <- player.df[order(player.df$APG, decreasing = TRUE), ]

# Selecting the top 10 players
top_10_assists_players <- top_assists_players[1:10, ]

# Creating a bar plot for the top 10 players by assists
ggplot(top_10_assists_players, aes(x = reorder(NAME, -APG), y = APG, fill = NAME)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Top 10 Players by Assists", x = "Player", y = "Assists Per Game") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Analysis: Eigenvalues

```{r}
p1_ess <- fviz_eig(pca, addlabels = TRUE, choice="eigenvalue",main ="Eigenvalues of the Explained Inertia")
p2_ess <- fviz_eig(pca, addlabels = TRUE, main="Proportion of Variances")
grid.arrange(p1_ess, p2_ess, nrow=1,name = "arrange")
```

## Analysis: What Makes Each PC

```{r}
# Contributions of variables to PC1
c1 <- fviz_contrib(pca, "var", axes=1)
c1
```

## Analysis: What Makes Each PC

```{r}
# Contributions of variables to PC2
c2 <- fviz_contrib(pca, "var", axes=2)
c2
```

## Analysis: What Makes Each PC

```{r}
# Contributions of variables to PC3
c3 <- fviz_contrib(pca, "var", axes=3)
c3
```

## Analysis: What Makes Each PC

```{r}
# Contributions of variables to PC4
c4 <- fviz_contrib(pca, "var", axes=4)
c4
```

```{r, include=FALSE, echo=FALSE}
# Identify numeric columns for histogram plotting
numeric_columns <- sapply(player.df, is.numeric)

# Get the column names of the numeric columns
numeric_column_names <- colnames(player.df)[numeric_columns]

# Create histograms for numeric columns
histograms <- lapply(numeric_column_names, function(col_name) {
  ggplot(player.df, aes_string(x = col_name)) +
    geom_histogram(binwidth = 5, fill = "skyblue", color = "black")
})

# Arrange the histograms in a grid
grid.arrange(grobs = histograms, ncol = 5)
```

```{r, include=FALSE, echo=FALSE}
names = player.df[,1]
grid.arrange(grobs = histograms, ncol = 4)
```



```{r, include=FALSE, echo=FALSE}
calculateScore = function(data) {
  return(sum((pca$rotation[, 1]*data)^2))
}
player.df$NAME[sort.int(apply(nba, 1, calculateScore), decreasing = T, 
                                           index.return = T)$ix[1:10]]

calculateScore = function(data) {
  return(sum((pca$rotation[, 1]*data)^2))
}
player.df$NAME[sort.int(apply(nba, 1, calculateScore), decreasing = T, 
                                           index.return = T)$ix[1:10]]

calculateScore = function(data) {
  return(sum((pca$rotation[, 2]*data)^2))
}
nba_data$GP[sort.int(apply(nba_data, 1, calculateScore), decreasing = T, index.return = T)$ix[1:10]]



head(get_pca_ind(pca)$contrib[,1])

head((pca$x[,1]^2)/(pca$sdev[1]^2))/dim(nba_data)[1]

names[order(get_pca_ind(pca)$contrib[,1],decreasing=T)][1:10]
```

## Analysis: Which Players Contribute Most

```{r}
names_z1 = names[order(get_pca_ind(pca)$contrib[,1],decreasing=T)]
fviz_contrib(pca, choice = "ind", axes = 1, top=20)+scale_x_discrete(labels=names_z1)
```

## Analysis: Which Players Contribute Most

```{r}
names_z2 = names[order(get_pca_ind(pca)$contrib[,2],decreasing=T)]
fviz_contrib(pca, choice = "ind", axes = 2, top=20)+scale_x_discrete(labels=names_z2)
```

## Analysis: Which Players Contribute Most

```{r}
names_z3 = names[order(get_pca_ind(pca)$contrib[,3],decreasing=T)]
fviz_contrib(pca, choice = "ind", axes = 3, top=20)+scale_x_discrete(labels=names_z3)
```

## Analysis: Which Players Contribute Most

```{r}
names_z4 = names[order(get_pca_ind(pca)$contrib[,4],decreasing=T)]
fviz_contrib(pca, choice = "ind", axes = 4, top=20)+scale_x_discrete(labels=names_z4)
```

## Analysis: Variables

```{r}
fviz_pca_var(pca, col.var = "cos2",
             gradient.cols = brewer.pal(3, "Set1"),
             repel = TRUE,
             title = "Cos 2 of the Variables NBA Stats - PCA")
```

## Analysis: Biplot

```{r,echo=FALSE}
fviz_pca_biplot(pca, col.var = "contrib", labelsize = 3, repel = TRUE, geom = "point", point.size = 2, point.color = "blue", title = "PCA Biplot Individuals and variables")
```

## Analysis: Players Overlay

```{r}
data.frame(z1=-pca$x[,1],z2=pca$x[,2]) %>% 
  ggplot(aes(z1,z2,label=names, color=nba$min)) + geom_point(size=0) +
  labs(title="Player ranking by minutes played on the PCA", x="Dim1 (33.4%)", 
       y="Dim2 (11.5%)") +
  theme_bw() + scale_color_gradient(low="red", high="darkblue")+
  theme(legend.position="bottom") + geom_text(size=2, hjust=0.6, vjust=0,
                                              check_overlap = TRUE)+
  guides(color = guide_legend(title = "Minutes played"))
```
